---
title: "Assignment 2: Wrapping up regression"
author: 
  - "Grace Stewart"
  - "DS303, SP26"
  - "Prof. Amber Camp"
date: today
format: html
toc: true
editor: visual
theme: spacelab
---

# Wrapping up linear and logistic regression^\*^

Assignment 2 covers linear and logistic regression models and includes many of the topics we have covered over the entire semester. You will be using `palmerpenguins` data, which includes a nice variety of continuous and categorical predictors.

To submit this assignment, render this file and save as a pdf. Commit & push this qmd to GitHub and upload the pdf to Canvas.

\*This assignment does not include mixed effects models, but you will be seeing that on another assignment.

## Load Packages

```{r, message = FALSE}
library(lme4)
library(tidyverse)
# install.packages("palmerpenguins") # install if needed
library(palmerpenguins)
```

## Load Data

Load the `penguins` data and examine it below (use `summary()`, etc.)

```{r}
penguins <- penguins

summary(penguins)
# ??palmerpenguins in terminal
```

## Question 1: Describe the data

What information is contained in this data set? Describe at least four variables (excluding `year`), including what they represent and their data type. Lastly, describe whether you think `year` would be a useful predictor in this data.

**Answer**:\
This data set includes information detailing various size measurements, sex, and home islands for Adelie, Chinstrap, and Gentoo penguins. For example, the penguins' measurements are noted through the following variables:

-   bill_length_mm: length of the penguin's bill measured in millimeters (mm), datatype = float

-   bill_depth_mm: depth of the penguin's bill measured in millimeters (mm), datatype = float

-   flipper_length_mm: length of the penguin's flipper measured in millimeters (mm), datatype = integer

-   body_mass_g: the weight of the penguin measured in grams (g), datatype = integer

I believe year would be a useful predictor in the data because year can help detect long-term changes in any possible evolutionary patterns found in the penguins. Additionally, it can help find patterns in trends over time such as climate change.

## Question 2: EDA

Explore your data visually. Create at least two visualizations that show the relationship between `flipper_length_mm` and its potential predictors.

```{r, warning = FALSE}

ggplot(penguins, aes(x=body_mass_g, y=flipper_length_mm, color=species)) +
  geom_point()
```

```{r}

ggplot(penguins, aes(x=species, y=flipper_length_mm)) +
  geom_boxplot()
```

## Question 3: Apply a linear regression

Fit a simple linear regression model predicting `flipper_length_mm` from `body_mass_g`.

```{r}

question3 <- lm(flipper_length_mm~body_mass_g, data=penguins)
summary(question3)
```

Interpret your model output, including the intercept and slope, in your own words below. Be sure to include a sentence explaining how `body_mass_g` impacts `flipper_length_mm` and whether or not the effect is significant.

**Answer:**\
This linear regression model predicts flipper length (mm) using body mass (g). The intercept (136.7) represents the length of the flipper when body mass is 0 grams. The slope of body mass is 1.528e-02 or 0.01528, meaning that for every 1 gram increase in body mass, flipper length increases by \~0.015mm. This shows that as body mass (g) increases, the flipper length (mm) will also increase. The p-value also shows that body_mass_g is a statistically significant in predicting flipper length. Additionally, the model can explain about 75% of variation found in flipper length.

## Question 4: Apply a multiple linear regression

Fit a linear regression model predicting `flipper_length_mm` from both `body_mass_g` and `bill_length_mm`. Interpret the slopes and intercept.

```{r}

question4 <- lm(flipper_length_mm ~ body_mass_g + bill_length_mm, data = penguins)
summary(question4)
```

Similar to Question 3, interpret the model output in your own words here:

**Answer:**\
This multiple linear regression model predicts flipper length (mm) using body mass (g) and bill length (mm). The intercept (122) represents the length of the flipper when body mass and bill length are 0. The slope of body mass is 1.305e-02 or 0.01305, meaning that for every 1 gram increase in body mass, flipper length increases by \~0.013mm. The slope of bill length is 5.492e-01 or 0.5492, meaning that for every 1 millimeter increase in bill length, flipper length increases by \~0.549mm. This shows that as body mass (g) and bill length (mm) increases, the flipper length (mm) will also increase. The p-values for both predictors show both variables are statistically significant in predicting flipper length. Additionally, the model can explain about 79% of variation found in flipper length.

## Question 5: Include an interaction

Fit a linear regression model predicting `flipper_length_mm` from `body_mass_g,` `bill_length_mm`, and the **interaction** of the two. Interpret the slopes and intercept.

```{r}

question5 <- lm(flipper_length_mm ~ body_mass_g * bill_length_mm, data = penguins)
summary(question5)
```

Interpret the model output in your own words below. If there was a change in the pattern of significance, try to explain the logic below as well.

**Answer:**\
This multiple linear regression model predicts flipper length (mm) using body mass (g) and bill length (mm) and their interaction. The intercept (170.6) represents the length of the flipper when body mass and bill length are 0. The coefficients are shown to be not statistically significant on their own, but with interaction is shown to be statistically significant, showing that the effect of body mass or bill length on flipper length rely on each other respectively. Additionally, the model can explain about 79% of variation found in flipper length.

## Question 6: Compare models

Compare the models you built in Questions 4 and 5 using `anova()`.

```{r}

anova(question4, question5)
```

-   Which is the better model? How do you know?

    -   **Answer:**\
        The interaction model is the better model. This is because the RSS decreased, representing a better fit in the model. Additionally, the p-value (0.002341) is statistically significant, showing that the interaction model is the better choice.

-   Is it possible to compare the models from Questions 3 and 5 using the same method? Why or why not?

    -   **Answer:**\
        Yes, it is possible to compare the models from Q3 to Q5 using ANVOA because it compares nested models, meaning one model is a subset of the interaction model. ANOVA tests whether adding additional predictors significantly improves model fit or not.

## Question 7: Categorical predictors

Build a linear model that includes a categorical predictor of your choice. It is fine to stick with dummy coding. Optional: apply a different coding scheme AND interpret the output correctly for +1 extra credit.

```{r}

#dummy coding
question7 <- lm(flipper_length_mm~species, data = penguins)
summary(question7)
```

-   What is the reference level of your categorical predictor?

    -   **Answer:**

        Adelie

-   What is your interpretation of this model output? Address all coefficients.

    -   **Answer:**

        This linear regression model predicts flipper length (mm) using species. The intercept (189.95) is the mean flipper length for the Adelie penguin species (reference level). Chinstrap penguins have flippers that are about 5.87mm longer than Adelie penguins whereas Gentoo penguins have flippers roughly 27.23mm longer than Adelie penguins. All coefficients are statistically significant. The model can explain about 78% of variation in flipper length, meaning species is a strong predictor of flipper length.

## Question 8: Relevel your categorical variable

Relevel your categorical variable so that a **different** level becomes the reference. Then, run the same model you did in Question 7 and interpret the output.

Relevel:

```{r}

#new reference = Gentoo
penguins$species <- relevel(penguins$species, ref = "Gentoo")
```

Apply same model from Question 7:

```{r}

question8 <- lm(flipper_length_mm ~ species, data = penguins)
summary(question8)
```

-   What is the new reference level of your categorical predictor?

    -   **Answer:**

        Gentoo

-   What is your interpretation of this new model output? Address all coefficients.

    -   **Answer:**

        This linear regression model predicts flipper length (mm) using species. The intercept (217.19) is the mean flipper length for the Gentoo penguin species (reference level). Chinstrap penguins have flippers that are about 21.36mm shorter than Gentoo penguins whereas Adelie penguins have flippers roughly 27.23mm shorter than Gentoo penguins. All coefficients are statistically significant. The model can explain about 78% of variation in flipper length, meaning species is a strong predictor of flipper length.

## Question 9: Fit a logistic regression

Apply a logistic regression. Include as many predictor variables as you'd like. Remember that your predicted outcome variable needs to be binary (or categorical with two levels).

Hint: You could use `sex` or create a binary variable of your own (e.g., Gentoo vs. non-Gentoo) to test your model.

```{r}

#check level
levels(penguins$sex)
###female, male

question9 <- glm(sex ~ body_mass_g + flipper_length_mm + bill_length_mm,
                 data = penguins,
                 family = binomial)
summary(question9)
```

What are your key takeaways from this model?

**Answer:**\
The key takeaway is that all predictors are statistically significant, this results in all meaningfully affecting the probability of the penguin being male. For example, for every 1 gram increase in body mass, the log odds of being male increase by 0.00296 and every 1 millimeter increase in bill length, the log odds of being male increase by 0.14589. On the other hand, with every 1 millimeter increase in flipper length, the log odds of being male *decrease* by 0.13486. This means that penguins that are heavier and have longer bills are more likely to be male, whereas penguins with longer flippers are more likely to be female.

## Question 10: Synthesize the information

Imagine you’re a biologist studying penguin populations. Which predictors do you think are most important to measure or record in the field to predict flipper length? Why?

**Answer:**\
If I were a biologist studying penguin populations, I feel the strongest predictors for flipper length would be species, body mass, and bill length. All three predictors account for a majority of variation in flipper length. For example species was able to explain roughly 78% of variation found in flipper length.

## Bonus: Stepwise Regression

Perform stepwise regression to find the best model for an outcome of your choice. You will likely encounter an error about missing values. Fixing that error and explaining your findings will earn you +2 extra credit. Show your work.

```{r}

#rm na values
penguins_clean <- na.omit(penguins)

#fitting full model
bonus <- lm(flipper_length_mm ~ ., data = penguins_clean)

#stepwise regression
stepwise <- step(bonus, direction = "both")
summary(stepwise)
```

According to this stepwise regression, explain how the final model was selected.

**Answer:**\
Stepwise regression selects the final model by removing predictors that do not meaningfully improve AIC (what measures how well a model fits the data). The only predictor that was removed was sex, meaning that it did not significantly improve the model's ability to predict flipper length.
