---
title: "Catching Legends"
subtitle: "An ML analysis on identifying legendary Pokemon"
author: "Grace Stewart"
institute: "Chaminade University"
date: 5/5/25
format: 
  revealjs:
    theme: moon
    transition: fade
    transition-speed: fast
editor: visual
---

## Background

Pokemon is a beloved franchise featuring various creatures that players can collect and battle

-   Each Pokemon has its own unique set of typings, statistics, and classifications

-   Some are categorized as "legendary" Pokemon

    -   Legendary Pokemon are considered far more rare and exclusive

## Research Question

What attribute can best predict a legendary Pokemon?

-   This analysis will explore whether or not Pokemon classified as "legendary" can be predicted through various key factors and attributes.

### Data

The data being used is from "The Complete Pokemon Dataset." The data was put together by user Rounak Banik and procured from Kaggle. It includes 802 Pokemon from Generations 1-7 (also known as the first 7 mainline games).

-   [https://www.kaggle.com/datasets/rounakbanik/pokemon](https://www.kaggle.com/datasets/rounakbanik/pokemon?utm_source=chatgpt.com)

```{r, echo=FALSE, output=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(plotly)
library(patchwork)
library(rpart)
library(rpart.plot)
library(Metrics)
```

## EDA {.scrollable}

```{r}
pkmn <- read.csv("data/pokemon.csv")

#dataset is alr clean so no need to search for NAs

pkmn$is_legendary <- factor(pkmn$is_legendary,
                            levels = c(0, 1),
                            labels = c("Not Legendary", "Legendary"))

ggplotly(ggplot(pkmn, aes(x = factor(is_legendary), fill = factor(is_legendary))) +
  geom_bar() +
  labs(
    title = "Number of legendary Pokemon",
    x = "Classification",
    y = "Count",
    legend = list(title = list(text = "Legendary Status"))
  ))
```

#### 

```{r}

ggplotly(ggplot(pkmn, aes(x = attack, y = defense, color = factor(is_legendary))) +
  geom_point() +
  labs(
    title = "Attack vs Defense by Legendary Status",
    color = "Legendary"
  ))
```

## Logistic Regression Model

```{r, echo=FALSE}

pkmn$capture_rate <- as.numeric(as.character(pkmn$capture_rate)) #changing capture_rate to numeric var instead of categorical

```

```{r, echo=TRUE, output=TRUE}

model <- glm(is_legendary ~ attack + defense + speed + hp + sp_attack + sp_defense + capture_rate,
             data = pkmn,
             family = binomial)

summary(model)
```

## Logistic Regression Model cont...

-   Strongest Predictor = Capture Rate
-   Statistically Significant:
    -   Speed & HP (p \< 0.001)
    -   Defense (p = 0.002)
    -   Special Attack (p = 0.003)

```{r, echo=TRUE, output=TRUE}

# calculating McFadden's R²
1 - model$deviance / model$null.deviance # = 0.600
```

-   Pseudo R²: 0.6003 or \~60%

## Decision Tree Model

```{r, echo=TRUE, output=TRUE}

tree_model <- rpart(
  is_legendary ~ attack + defense + speed + hp + sp_attack + sp_defense + capture_rate, # predict Species from all other variables
  data = pkmn,
  method = "class") # using "class" for classification,

summary(tree_model)
```

## Decision Tree cont...

```{r}
#| fig-width: 8
#| fig-height: 5

rpart.plot(
  tree_model,
  type = 4,
  extra = 104, # show class probabilities and % of observations
  fallen.leaves = TRUE,
  main = "Decision Tree: Legendary Pokemon")

```

## Comparisons

::::: columns
::: {.column width="50%"}
Logistic Regression

-   Capture rate is strongest predictor

-   Some stats (HP, Speed, Defense, Sp. Attack) are significant

-   Strong overall fit (60% pseudo R²)

    -   Potential overfitting
:::

::: {.column width="50%"}
Decision Tree

-   Capture rate is what cleanly splits the data the most

    -   Pokemon with \<20 capture rate are highly likely to be legendary

-   Model strongly classifies majority as non-legendary first
:::
:::::
