---
title: "DS303 Final Project"
author: 
  - "Grace Stewart"
  - "DS303, SP26"
  - "Prof. Amber Camp"
date: 4/24/26
format: html
toc: true
editor: visual
theme: spacelab
---

## Load packages

```{r, message = FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(plotly)
library(patchwork)
library(rpart)
library(rpart.plot)
library(Metrics)
```

## Load data

The date being used is from "The Complete Pokemon Dataset." The data was put together by user Rounak Banik and procured from Kaggle.

```{r, message= FALSE}
pkmn <- read.csv("data/pokemon.csv")

summary(pkmn)
```

## Exploratory Data Analysis (EDA)

```{r}

#comparing atk and sp.atk
ggplotly(ggplot(pkmn, aes(x=attack, y=sp_attack, color=type1)) +
  geom_point(alpha = 0.5))

#how many in each generation
ggplotly(ggplot(pkmn, aes(x=legendary, color=legendary)) +
  geom_bar())

#comparing height and weight
ggplotly(ggplot(pkmn, aes(x=height_m, y=weight_kg, color=generation)) +
  geom_point(alpha = 0.5))
```

## Modeling

```{r}

pkmn$capture_rate <- as.numeric(as.character(pkmn$capture_rate)) #changing capture_rate to numeric var instead of categorical


model <- glm(is_legendary ~ attack + defense + speed + hp + sp_attack + sp_defense + capture_rate,
             data = pkmn,
             family = binomial)

summary(model)
```

-   capture rate is strongest predictor

    -   legendaries are harder to catch likely cause rarer

-   spd and hp matter most compared to other stats

    -   usually faster and bulkier in gameplay

```{r}

# calculating McFadden's R²
1 - model$deviance / model$null.deviance # = 0.600
```

-   Pseudo R²: 0.6003 or \~60%
    -   model can explain 60% of deviance, strong fit but could be overfitted

```{r}

tree_model <- rpart(
  is_legendary ~ attack + defense + speed + hp + sp_attack + sp_defense + capture_rate, # predict Species from all other variables
  data = pkmn,
  method = "class") # using "class" for classification,

summary(tree_model)
```

-   capture rate = main predictor

    -   root split/very decisive

    -   capture rate \< 20 = identified 90% of legendary pkm

```{r}

#| fig-width: 8
#| fig-height: 5

rpart.plot(
  tree_model,
  type = 4,
  extra = 104, # show class probabilities and % of observations
  fallen.leaves = TRUE,
  main = "Decision Tree: Legendary Pokemon")
```

## Comparisons

```{r}

library(corrplot)

vars <- pkmn[, c("attack", "defense", "speed", "hp", "sp_attack", "sp_defense", "capture_rate")]

m <- cor(vars, use = "complete.obs")
corrplot(m)
corrplot(abs(m))
```

-   negative correlation for capture rate

    -   lower capture rate for stronger pokemon (legendary)

-   positive correlation for mostly everything else

-   sp.atk/def slightly stronger

-   no multicollinearity

```{r}

# decision tree confusion matrix
pred <- predict(tree_model, pkmn, type = "class")
table(Predicted = pred, Actual = pkmn$is_legendary)
```

-   (723+63)/801= \~98% accuracy

-   63/70 legendaries

-   misclassified 15 pkmn

```{r}

##test train split
# split data
set.seed(123)
train_idx <- sample(1:nrow(pkmn), 0.8 * nrow(pkmn))
train <- pkmn[train_idx, ]
test  <- pkmn[-train_idx, ]

tree_model <- rpart(is_legendary ~ attack + defense + speed + hp + sp_attack + sp_defense + capture_rate, data = train, method = "class")

pred_test <- predict(tree_model, test, type = "class")
confusionMatrix(pred_test, as.factor(test$is_legendary), positive = "1")
```

-   98% accuracy

-   87% sensitivity

-   99% specificity
