---
title: "Coding Categorical Variables"
author: 
  - "Student Name"
  - "DS303, SP25"
  - "Prof. Amber Camp"
date: 2/19/25
format: html
editor: visual
theme: spacelab
---

## Coding Categorical Predictors

Working with categorical predictors is an important aspect of building effective statistical models. Categorical predictors, also known as factors, represent qualitative data that can be organized into distinct groups, such as gender, geographic location, or product type.

In R, these predictors can be integrated into various modeling techniques, helping us identify relationships and make predictions based on different *levels* of these variables.

## UCLA Reference

There are many resources out there, but this is one I go back to again and again. I find it really easy to understand and to also use as a reference:

[**R LIBRARY CONTRAST CODING SYSTEMS FOR CATEGORICAL VARIABLES**](https://stats.oarc.ucla.edu/r/library/r-library-contrast-coding-systems-for-categorical-variables/)

## Set up your environment

### Load packages

```{r, echo = FALSE, message = FALSE}
library(lme4)
library(tidyverse)
```

### Load Data

Following the UCLA resource, we will use data frame called `hsb2` and focusing on the categorical variable `race`, which has four levels (1 = Hispanic, 2 = Asian, 3 = African American and 4 = Caucasian). We will use `write` as our dependent variable.

[Here](https://www.openintro.org/data/index.php?data=hsb2) is a brief description of `hsb2` data (but note that they have filled in some of their values a bit differently than we have below).

```{r}
# read in data from URL
hsb2 <- read.table('https://stats.idre.ucla.edu/stat/data/hsb2.csv', header=T, sep=",")

skim <- skimr::skim(hsb2)
summary(as.factor(hsb2$race))
```

Let's create `race.f`, which is the `race` data re-labeled and also turned into factors.

```{r}
hsb2$race.f <- factor(hsb2$race, labels=c("Hispanic", "Asian", "African-Am", "Caucasian"))
```

### Inspect your data

Inspect your data below using `skimr`, `summary`, and `ggplot2`. You should also check out [this resource](https://www.openintro.org/data/index.php?data=hsb2), which describes the dataset.

```{r}
skim <- skimr::skim(hsb2)
View(skim)

summary(hsb2)
summary(hsb2$race)
summary(hsb2$race.f)

ggplot(hsb2, aes(x = race.f, y = write)) +
  geom_boxplot()

ggplot(hsb2, aes(x = race.f, y = write)) +
  geom_jitter(width = 0.2, height = 0) 

```

## Coding Schemes

There are several different coding schemes, and you can also generate your own scheme for your own purposes. For our class, we will focus on **Dummy**, **Simple**, **Deviation**, and **Helmert** Coding.

We've already seen dummy coding, but it is good to see it again.

## Class activity

Split into four groups. Explore your assigned coding scheme and answer the below five questions. Then, apply your coding scheme to the `hsb2` data and walk us through its interpretation.

## Dummy Coding (a.k.a. Treatment Coding)

1.  What comparisons are made with this coding scheme? How does this coding scheme affect the coefficients in regression models? (What do the coefficients represnt? What is the reference level?)\
    \

2.  When is this coding scheme preferred? What types of data or analysis scenarios make this coding method particularly useful?\
    \

3.  What are the advantages of this coding method? How does it simplify interpretation or enhance model performance?\
    \

4.  What are the limitations of this coding method? Are there situations where this method may lead to misleading results or interpretations?\
    \
    \

### Applying Dummy Coding

```{r}
# assign dummy coding
contrasts(hsb2$race.f)

# build and summarize your regression
dummy <- lm(write ~ race.f, data = hsb2)
summary(dummy)
```

Summarize these findings. What is your reference group, and how do the other groups compare with regards to their writing scores? What does the intercept value represent?\
\
(write here)\

## Simple Coding

1.  What comparisons are made with this coding scheme? How does this coding scheme affect the coefficients in regression models? (What do the coefficients represnt? What is the reference level?)\

2.  When is this coding scheme preferred? What types of data or analysis scenarios make this coding method particularly useful?\

3.  What are the advantages of this coding method? How does it simplify interpretation or enhance model performance?\

4.  What are the limitations of this coding method? Are there situations where this method may lead to misleading results or interpretations?\

### Applying Simple Coding

```{r}
# assign simple coding
#creating the contrast matrix manually by modifying the dummy coding scheme
c <- contr.treatment(4)
my.coding <- matrix(rep(1/4, 12), ncol=3)
my.simple <- c-my.coding
my.simple


contrasts(hsb2$race.f) <- my.simple
summary(lm(write ~ race.f, hsb2))


# build and summarize your regression
simple <- lm(write ~ race.f, data = hsb2)
summary(simple)

?contr.treatment
```

Summarize these findings. What is your reference group, and how do the other groups compare with regards to their writing scores? What does the intercept value represent?\
\
(write here)\

## Deviation Coding (a.k.a. Effect Coding)

(Grace)

1.  What comparisons are made with this coding scheme? How does this coding scheme affect the coefficients in regression models? (What do the coefficients represnt? What is the reference level?)

    1.  Deviation coding is used in regression analysis to compare the mean of a specific category (level) to the average of all group means (grand mean). It essentially compares how the group mean deviates from the overall mean.

    2.  This coding scheme affects the coefficients by having them represent the difference between a specific group's mean and the overall grand mean of the dependent variable. There is no real reference level, rather the excluded group is represented by the negative sum of coefficient of the included groups.\

2.  When is this coding scheme preferred? What types of data or analysis scenarios make this coding method particularly useful?

    1.  This coding scheme is preferred for balanced experimental designs, factorial designs, when there is no natural reference group or when examining how much each group deviates from the overall mean.

    2.  Some data science/analysis scenarios in which this method is NOT useful is when there is a clear, meaningful control group or you are directly comparing something to a baseline.

        1.  Clinical trials could go either way:

            1.  Yes: Comparing the effects of 2 medications

            2.  No: Treatment vs. Placebo

3.  What are the advantages of this coding method? How does it simplify interpretation or enhance model performance?

    1.  Some advantages of deviation coding:

        1.  The intercept is meaningful: because it is the grand mean which represents the overall average outcome across all conditions

        2.  Coefficients will show above or below average: makes interpretations simple (positive = above average, negation = below average)

        3.  There is no arbitrary reference group: no group is considered the baseline which makes all groups "equal". Helps when there are no natural control group, symmetric comparisons, etc.

    2.  It does not change the overall model fit. It only changes how coefficients are interpreted and how you can frame your hypothesis.

4.  What are the limitations of this coding method? Are there situations where this method may lead to misleading results or interpretations?

    1.  This coding scheme is not ideal when there is a true control group because it can obscure the main comparisons as it is tied to the grand mean instead. Also when group sizes are unequal (unbalanced design), the grand mean is a weighted mean, meaning larger groups will have more influence.

    2.  This can be misleading when there is a baseline group or group sizes are highly unequal.

### Applying Deviation Coding

```{r}
# assign deviation coding
contr.sum(4)


#assigning the deviation contrasts to race.f
contrasts(hsb2$race.f) = contr.sum(4)
#the regression
summary(lm(write ~ race.f, hsb2))

hsb2$race.f <- relevel(hsb2$race.f, ref = "Asian")


# build and summarize your regression
dev <- lm(write ~ race.f, data = hsb2)
summary(dev)

# Get the mean of the fourth group (e.g., "Caucasian")
mean_caucasian <- mean(subset(hsb2, race.f == "Caucasian")$write)

# Compare it to the grand mean (intercept value)
grand_mean <- coef(lm(write ~ race.f, data = hsb2))["(Intercept)"]


means_tapply <- tapply(hsb2$write, hsb2$race.f, mean)
print(means_tapply)

m_o_m <- (46.45833+58+48.2+54.05517)/4 #mom = mean of means
g_m <- mean(hsb2$write) #gm = grand mean

summary(hsb2$race.f)
```

Summarize these findings. Unlike dummy and simple coding, there is no traditional reference level in deviation coding. In this context, how do the groups compare regarding their writing scores? What does the intercept value represent?

-   Intercept = 51.68 (grand mean)

-   Each coefficient shows how each group's mean differs from this average. Asian students score significantly above the grand mean whereas Hispanic and African-American students score below. This shows a symmetric comparison in the sense that all the groups are evaluated relative to the overall mean how much they deviate from that number.\

## Helmert Coding

1.  What comparisons are made with this coding scheme? How does this coding scheme affect the coefficients in regression models? (What do the coefficients represnt? What is the reference level?)\
    \

2.  When is this coding scheme preferred? What types of data or analysis scenarios make this coding method particularly useful?\
    \

3.  What are the advantages of this coding method? How does it simplify interpretation or enhance model performance?\
    \

4.  What are the limitations of this coding method? Are there situations where this method may lead to misleading results or interpretations?\
    \
    \
    \

### Applying Helmert Coding

```{r}
# assign Helmert coding
#helmert for factor variable with 4 levels
my.helmert = matrix(c(3/4, -1/4, -1/4, -1/4, 0, 2/3, -1/3, -1/3, 0, 0, 1/
	2, -1/2), ncol = 3)
my.helmert
      

#assigning the new helmert coding to race.f
contrasts(hsb2$race.f) = my.helmert
contrasts(hsb2$race.f)

hsb2$race.f <- relevel(hsb2$race.f, ref = "Caucasian")
contrasts(hsb2$race.f) = my.helmert

# build and summarize your regression
helmert <- lm(write ~ race.f, data = hsb2)
summary(helmert)
```

Summarize these findings. In Helmert coding, each level is compared to the mean of subsequent levels, rather than having a traditional reference group. In this context, how do the groups compare regarding their writing scores? What does the intercept value represent?\
\
Answer: Each group's writing score is compared to the mean of the subsequent levels, instead of using a traditional reference group. The intercept value represents the average writing score across all levels before any comparisons are made. The coefficients show how each group’s writing score differs from the average of the following groups.

## For next class: Releveling a predictor

We will do this together as a class. You don't need to work ahead, but if you wanted to explore releveling, here you go.

Releveling refers to the process of changing the reference level of a factor or categorical variable. This allows us to examine how different groups compare to a new reference category. By selecting a different reference level, we can gain new insights into the effects of categorical predictors in our analyses.

Releveling is applicable to Dummy, Simple, and Helmert coding, but NOT deviation coding. Why?

### To relevel:

You can use the below to define a reference level, or you can assign new values in your contrasts matrix.

```{r}
hsb2$race.f <- relevel(hsb2$race.f, ref = "Asian")
```
