---
title: "Multiple Linear Regression: Interaction Terms"
author: 
  - "Grace Stewart"
  - "DS303, SP25"
  - "Prof. Amber Camp"
date: 2/4/26
format: html
editor: visual
theme: spacelab
---

## Interactions in Multiple Regression

When we add an interaction term, we're asking:

Does the relationship between *x₁* and *y* *depend on* the value of *x₂*?

### Setup

```{r}
library(ISLR2)
library(lme4)
library(tidyverse)

Boston <- Boston
```

## The basic idea

Without interaction:

-   The effect of rooms on home value is **the same** whether the neighborhood has high or low crime

With interaction:

-   The effect of rooms on home value **changes** depending on crime levels

-   Maybe extra rooms add more value in low-crime areas than high-crime areas

## The math

**Without interaction:** *medv* = *b₀* + *b₁*\**rm* + *b₂*\*crim

```{r}
model_no_int <- lm(medv ~ rm + crim, data = Boston)
summary(model_no_int)
```

**With interaction:** *medv* = *b₀* + *b₁rm + b₂crim + b₃(rm*\**crim*)

```{r}
model_int <- lm(medv ~ rm + crim + rm:crim, data = Boston)

# or use the shortcut:
model_int <- lm(medv ~ rm * crim, data = Boston)

summary(model_int)
```

That last term (*b₃(rm*\**crim*)) is the interaction.

## Interpreting interactions

This is where it gets tricky.

Here's what changes when you add an interaction term:

**Without interaction**, the coefficient for `rm` means:

-   "For every additional room, median home value increases by \$X, holding `crim` constant"

-   This effect is **the same** regardless of the value of `crim`

**With interaction**, the coefficient for `rm` means:

-   "For every additional room, median home value increases by \$X, **when crim = 0**"

-   This effect **changes** depending on the value of crim

-   The interaction term tells you **how much** the effect of rm changes per unit of crim

The main effects are still interpretable, but they now represent **conditional effects** (the effect when the other variable equals zero), rather than a single constant relationship.

```{r}
# same thing, just printed again
model_int <- lm(medv ~ rm * crim, data = Boston)
summary(model_int)
```

-   **Intercept (*b₀* = -37.26):** The predicted median home value when a home has zero rooms and `crim` is zero is -\$37,257 (not practically meaningful since you can't have zero rooms...)

-   **`rm` (*b₁* = 9.65):** For every additional room, median home value increases by \$9,651, **when `crim` = 0** (holding `crim` constant at zero)

-   **`crim` (*b₂* = 1.46):** For every one-unit increase in `crim`, median home value increases by \$1,463, **when `rm` = 0** (holding rooms constant at zero... counterintuitive, but a good sanity check)

-   **`rm:crim` (*b₃* = -0.28):** For every one-unit increase in `crim`, the effect of an additional room on median home value **decreases** by \$288

    Or equivalently: For every additional room, the effect of `crim` on median home value decreases by \$288.

NOTE: The interaction interpretation is the key one here. The negative interaction means that rooms are worth MORE in low-crime neighborhoods and worth LESS in high-crime neighborhoods.

### Understanding what the interaction means

The interaction coefficient tells us: **"How much does the effect of rooms change for each one-unit increase in `crim`?"**

Or equivalently: **"How much does the effect of `crim` change for each additional room?"**

**Important**: When an interaction is present, the effect of `rm` is no longer constant. It depends on `crim`, and vice versa.

## When to include interactions

Add interaction terms when you have theoretical reasons to believe the effect of one variable depends on another:

-   **Urban vs. rural**: Does education affect income differently in cities vs. countryside?

-   **Treatment dosage × patient age**: Do medications work differently for different age groups?

-   **Marketing spend × season**: Does advertising work better during holidays?

Don't just throw in interactions randomly. The model gets complicated fast and harder to interpret!

# Interaction Plots

### `interaction` method

The three lines represent the rooms-value relationship at different crime levels (mean, +1 SD, -1 SD). The spacing between the lines shows the interaction effect - as crime increases, the positive relationship between rooms and home value weakens. The confidence bands (shaded areas) show we're quite certain about these different slopes.

```{r}
# install.library("interactions")
library(interactions)

# fit your model
model_int <- lm(medv ~ rm * crim, data = Boston)

# create interaction plot
interact_plot(model_int, 
              pred = rm,           # x-axis variable
              modx = crim,         # moderator variable
              interval = TRUE,     # show confidence intervals
              int.width = 0.95)    # 95% CI

```

### `ggplot` method

The relationship between number of rooms and home value differs across crime levels. In low-crime neighborhoods (green line), each additional room is associated with a steeper increase in home value compared to high-crime neighborhoods (red line), where the slope is much flatter. This visual pattern confirms our interaction term: the effect of rooms on home value depends on the crime level.

```{r}
# create groups based on crime levels (low, medium, high)
Boston <- Boston %>%
  mutate(crim_level = case_when(
    crim < quantile(crim, 0.33) ~ "Low Crime",
    crim < quantile(crim, 0.67) ~ "Medium Crime",
    TRUE ~ "High Crime"))

# plot with separate lines
ggplot(Boston, aes(x = rm, y = medv, color = crim_level)) +
  geom_point(alpha = 0.3) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(title = "Interaction: Rooms × Crime Level",
       x = "Number of Rooms",
       y = "Median Home Value ($1000s)",
       color = "Crime Level") +
  theme_minimal()

```

# Practice

Build a model using `Boston` data that contains an interaction term and interpret the results below.

```{r}

model_practice <- lm(medv ~ rm * age, data = Boston)
summary(model_practice)
```

Interpretation:

-   Intercept: ***b₀*** **=** -59.59286, The predicted medv when a home has 0 rooms and the proportion of owner-occupied unit built prior to 1940 is 0, is -\$59,592.86.

-   Main effect 1: rm (***b₁*** **=** 13.74747), For every additional room, medv increases by \$13,747.47, when age=0 (holding age at a constant 0).

-   Main effect 2: age (***b₂*** **=** 0.38153), For every one unit increase in age, medv increases by \$381.53, when rm=0 (holding rm at a constant 0).

-   Interaction term: rm:age (***b₃*** **=** -0.07141), For every one unit increase in age, the effect of an additional room on medv decreases by \$71.41.

    -   For every additional room, the effect of age on medv decreases by \$71.41.

-   Which terms are statistically significant at α = 0.05?

    -   All terms were statistically significant at α = 0.05.

-   Build a second model with the same terms but without the interaction. Compare the R² values between your two models. Does adding the interaction improve the model?

```{r}

model_practice2 <- lm(medv ~ rm + age, data = Boston)
summary(model_practice2)
```

Try visualizing the interaction (use the `interaction` package)

```{r}

# create interaction plot
interact_plot(model_practice, 
              pred = rm,           # x-axis variable
              modx = age,         # moderator variable
              interval = TRUE,     # show confidence intervals
              int.width = 0.95)    # 95% CI
```

# Bonus (scary stuff)

You can have a three-way interaction! It's kind of mind-bending...

```{r}
model_three <- lm(medv ~ rm * crim * lstat, data = Boston)
summary(model_three)
```

That last term, `rm:crim:lstat`, would be interpreted like this:

-   For every one-unit increase in `lstat`, the interaction effect between `rm` and `crim` changes by 0.010

-   *"The extent to which crime moderates the effect of rooms on home value itself depends on the socioeconomic status of the neighborhood."*

This is really difficult to interpret meaningfully. Let's avoid this for now.
